---
title: "EDA projects in R"
author: "Musyoka James"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Load libraries

```{r} 
library(tidyverse)
library(dplyr)
library(ggplot2)

```

#Economic Trend analysis
```{r } 

head(economics)

tail(economics)

summary(economics)

str(economics)

#Plotting unemployment rate over time

economics |> 
  ggplot(aes(date, unemploy))+
  geom_line()+
  labs(title = "Unemployment rate over time", x = "Year", y = "Unemplyoyed")+
  annotate("text", x = as.Date("2008-01-01"), y = 10500, label = "2008 Recession", color = "red")+
  theme_bw()
```


The Number of unemployed people keeps increasing as time goes on

The highest peak of unemployment was in 2010

From the early 1960s, unemployment shows a generally fluctuating trend with several sharp rises and falls, indicating periods of economic downturn and recovery

Significant spikes in unemployment are seen around the early 1980s and late 2000s, possibly corresponding to known recessions.

I have annotated the 2008 Recession where the number of unemployment cases skyrocketed due to harsh economic times and layoffs.


## Explore housing market
```{r} 
#View(txhousing)

str(txhousing)

head(txhousing)

summary(txhousing)

#We only want to look at the median home price in Austin

#Hence we filter
txhousing |> 
  filter(city == "Austin") |> 
  group_by(year) |> 
  summarise(median = median(median)) |> 
  ggplot(aes(year, median))+
  geom_line() +
  geom_point(aes(size = 2, alpha = 0.5))+
  geom_smooth(se = F)+
  theme_classic()

```


Between 2000 and 2015, the median home price in Austin showed a consistent upward trend, indicating a growing housing market. A noticeable dip occurred around 2008–2009, possibly due to the global financial crisis. Since 2010, the prices have risen more rapidly, reflecting increased demand and urban growth in Austin.


##Explore Storm Intensities
```{r}
#View(storms)

str(storms)

head(storms)

tail(storms)

summary(storms)

#Identify the most intense hurricanes
storms |> 
  filter(status == "hurricane") |> 
  arrange(desc(wind))


#Compare wind distributions across status
storms |> 
  ggplot(aes(x = status, y = wind))+
  geom_boxplot()+
  theme_classic()+
  coord_flip()

```



#Global life expectancy and GDP
#Compare life expectancy vs gdp per capita for 2007

```{r}
library(gapminder)
#View(gapminder)

head(gapminder)

str(gapminder)

summary(gapminder)

gapminder |> 
  filter(year == 2007) |> 
  ggplot(aes(gdpPercap, lifeExp))+
  geom_point(aes(colour = continent), size = 2)+
  geom_smooth(se = F, method = lm)+
  scale_x_log10()+
  theme_bw()+
  labs(title = "Life Exp VS GDP per Capita in 2007",
        x = "GDP per Capita (log scale)",
        y = "Life Expectancy")


```


In 2007, there was a positive linear relationship between GDP per capita and life expectancy across countries. As GDP per capita increased (on a log scale), life expectancy tended to rise. This suggests that wealthier countries generally had longer-lived populations.

There may still be variation by continent (shown by color), but the overall global trend is upward.


#Flight delay analysis
```{r } 
#Load the required packages
library(nycflights13)

#View(flights)
str(flights)

head(flights)

tail(flights)

summary(flights)

#We need the airlines dataset in this project.
#We are going to join the two datasets 

summary(airlines)

#Find airlines with the worst average delay
flights |> 
  select(dep_delay, arr_delay, carrier, month) |> 
  drop_na() |> 
  left_join(airlines) |> 
  group_by(name) |> 
  summarise(avg_dep_delay = mean(dep_delay)) |> 
  arrange(desc(avg_dep_delay)) 

#Analyze delay trends over months
flights |> 
  select(dep_delay, arr_delay, month) |> 
  drop_na() |> 
  group_by(month) |> 
  summarise(avg_arr_delay = mean(arr_delay)) |> 
  ggplot(aes(month, avg_arr_delay))+
  geom_point()+
  geom_line()+
  theme_bw()+
  labs(title = "Average Arrival Delay Through The Months", x = "Month", y = "Average Arrival Delay")
```
Arrival delays vary throughout the year, showing a general pattern of increases and decreases across months. Certain months e.g. the 6th and 7th months experience higher average delays, while others (e.g., the 9th month) have lower delays- it is in the negatives.

This trend may reflect seasonal factors like bad weather, holiday travel spikes, or air traffic congestion during specific times of the year.

Arrival delays showed noticeable variation across the year. The highest delays occurred in June and July, suggesting possible seasonal congestion or weather-related issues during summer travel.

In contrast, September and October had the lowest average delays, even dipping into negative values, indicating that flights tended to arrive earlier than scheduled in those months.

After October, delays rose sharply, possibly due to increased holiday travel or weather disruptions toward the end of the year.






